---
title: "Causes of Death In the US"
author: "Dallin"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---
#Install necessary packages in order to analyze data.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("dplyr")
library(dplyr)
library(stringr)
library(readr)
library(ggplot2)
```
#Import data and filter data
```{r, s}
#Data from 2000 to 2017
library(readr)
options(readr.show_col_types = FALSE)
death.total.table <- read_csv("~/Downloads/NCHS_-_Leading_Causes_of_Death__United_States (3).csv")
death.total <- data.frame(death.total.table)
#View(death.total)
#Filter data by causes of death pertaining to study
filtered.death.total <- death.total %>% 
  group_by(Year) %>% 
  filter(State=="United States", Year > 1999, Cause.Name != "All causes", Cause.Name != "Unintentional injuries", Cause.Name != "Suicide") %>% 
  subset(select = c(-Age.adjusted.Death.Rate,-X113.Cause.Name))
#View(filtered.death.total)
#add rows for the upcoming years
death.total.table.2 <- read_csv("~/Downloads/Final Project LR data 2018-2020 - Sheet1.csv")
death.total.2 <- data.frame(death.total.table.2)
#View(death.total.2)
combined.death.total <- data.frame(rbind(filtered.death.total, death.total.2))
#View(combined.death.total)
#mutate the population so that it is per 100,000 people
#population per year for rate question
population.year <- c(rep(325122128, times = 8), rep(323071755, times = 8), rep(320738994, times = 8), rep(318386329, times = 8), rep(316059947, times = 8), rep(313877662, times = 8), rep(311583481, times = 8), rep(309327143, times = 8), rep(306771529, times = 8), rep(304093966, times = 8), rep(301231207, times = 8), rep(298379912, times = 8), rep(295516599, times = 8), rep(292805298, times = 8), rep(290107933, times = 8), rep(287625193, times = 8), rep(284968955, times = 8), rep(282162411, times = 8), rep(326838199, times = 8), rep(328329953, times = 8), rep(331501080, times = 8))
combined.death.total %>% 
  group_by(Year)
combined.death.total$population <- population.year
ordered.combined.death.total <- combined.death.total[order(combined.death.total$Year),]
#View(ordered.combined.death.total)
final.data <- ordered.combined.death.total %>% 
  group_by(Year) %>% 
  subset(select = -State) %>% 
  mutate(rate = (Deaths/population)*100000)
rep_string = c("Alzheimer's Disease"="Alzheimer's disease", "Heart Disease"="Heart disease", "Kidney Disease"="Kidney disease")
final.data$Cause.Name <- str_replace_all(final.data$Cause.Name, rep_string)
#Import Data on Health Expenditures
health.expenditures <- data.frame(read_csv("~/Downloads/costs - Sheet1.csv"))
#View(health.expenditures)
health_expenses <- c(rep(1365.7, times = 8), rep(1483.1, times = 8), rep(1630.6, times = 8), rep(1769.9, times = 8), rep(1894.1, times = 8), rep(2025.9, times = 8), rep(2164.4, times = 8), rep(2305, times = 8), rep(2402, times = 8), rep(2492.5, times = 8), rep(2589.4, times = 8), rep(2676.2, times = 8), rep(2782.8, times = 8), rep(2855.8, times = 8), rep(3001.4, times = 8), rep(3163.6, times = 8), rep(3305.6, times = 8), rep(3446.5, times = 8), rep(3604.5, times = 8), rep(3759.1, times = 8), rep(4124, times = 8))
#final.data is the data, all prepared for work
final.data$health <- health_expenses
View(final.data)
print(final.data)
```
#Main Plot for this Statistical Analysis
```{r}
#Plot of the eight causes of death. 
#X-axis is from the years 2000 to 2020, which is one of the independent variables. 
#Y-axis is measuring the mortality rate of each different disease. These are the multiple independent variables that I am analyzing.
ggplot(final.data, aes(Year, rate, color = Cause.Name)) +
  ggtitle("Year vs Mortality Rate for Leading Causes of Death") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  scale_color_discrete("Cause of Death") +
  xlab("Year") +
  ylab("Mortality Rate per 100k") 
ggplot(final.data, aes(health, rate, color = Cause.Name)) +
  ggtitle("Health Expenditure vs Mortality Rate for Leading Causes of Death") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  scale_color_discrete("Cause of Death") +
  xlab("Health Expenditure in Billions") +
  ylab("Mortality Rate per 100k") 
```
#Different Important Numbers for analysis: 
```{r}
#Pearson's correlation coefficient: I'm looking for which diseases demonstrate a positive or negative correlation as time with the time in years.
#heart disease
heart_disease <- filter(final.data, Cause.Name == "Heart disease")
#View(heart_disease)
print("Pearson coefficient: Year vs Mortality Rate of Heart Disease")
cor(heart_disease$Year, heart_disease$rate)

cancer_disease <- filter(final.data, Cause.Name == "Cancer")
print("Pearson coefficient: Year vs Mortality Rate of Cancer")
cor(cancer_disease$Year, cancer_disease$rate)

stroke_disease <- filter(final.data, Cause.Name == "Stroke")
print("Pearson coefficient: Year vs Mortality Rate of Stroke")
cor(stroke_disease$Year, stroke_disease$rate)

clrd_disease <- filter(final.data, Cause.Name == "CLRD")
print("Pearson coefficient: Year vs Mortality Rate of CLRD")
cor(clrd_disease$Year, clrd_disease$rate)

diabetes_disease <- filter(final.data, Cause.Name == "Diabetes")
print("Pearson coefficient: Year vs Diabetes")
cor(diabetes_disease$Year, diabetes_disease$rate)

alzheimers_disease <- filter(final.data, Cause.Name == "Alzheimer's disease")
print("Pearson coefficient: Year vs Mortality Rate of Alzheimer's Disease")
cor(alzheimers_disease$Year, alzheimers_disease$rate)

kidney_disease <- filter(final.data, Cause.Name == "Kidney disease")
print("Pearson coefficient: Year vs Mortality Rate of Kidney Disease")
cor(kidney_disease$Year, kidney_disease$rate)

flu.pneu_disease <- filter(final.data, Cause.Name == "Influenza and pneumonia")
print("Pearson coefficient: Year vs Mortality Rate of Flu and Pneumonia")
cor(flu.pneu_disease$Year, flu.pneu_disease$rate)

#Pearson's correlation coefficient: I'm looking for which diseases demonstrate a positive or negative correlation with health expenditures
#heart disease
heart_disease <- filter(final.data, Cause.Name == "Heart disease")
#View(heart_disease)
print("Pearson coefficient: Health Expenditures vs Mortality Rate of Heart Disease")
cor(heart_disease$health, heart_disease$rate)

cancer_disease <- filter(final.data, Cause.Name == "Cancer")
print("Pearson coefficient: Health Expenditures vs Mortality Rate of Cancer")
cor(cancer_disease$health, cancer_disease$rate)

stroke_disease <- filter(final.data, Cause.Name == "Stroke")
print("Pearson coefficient: Health Expenditures vs Mortality Rate of Stroke")
cor(stroke_disease$health, stroke_disease$rate)

clrd_disease <- filter(final.data, Cause.Name == "CLRD")
print("Pearson coefficient: Health Expenditures vs Mortality Rate of CLRD")
cor(clrd_disease$health, clrd_disease$rate)

diabetes_disease <- filter(final.data, Cause.Name == "Diabetes")
print("Pearson coefficient: Health Expenditures vs Diabetes")
cor(diabetes_disease$health, diabetes_disease$rate)

alzheimers_disease <- filter(final.data, Cause.Name == "Alzheimer's disease")
print("Pearson coefficient: Health Expenditures vs Mortality Rate of Alzheimer's Disease")
cor(alzheimers_disease$health, alzheimers_disease$rate)

kidney_disease <- filter(final.data, Cause.Name == "Kidney disease")
print("Pearson coefficient: Health Expenditures vs Mortality Rate of Kidney Disease")
cor(kidney_disease$health, kidney_disease$rate)

flu.pneu_disease <- filter(final.data, Cause.Name == "Influenza and pneumonia")
print("Pearson coefficient: Health Expenditures vs Mortality Rate of Flu and Pneumonia")
cor(flu.pneu_disease$health, flu.pneu_disease$rate)

```
#Individual Causes of Death plotted against time in years. Each graph shows a straight line which represents the linear trend of the data and a shadowed area which represents the confidence interval of 95%
```{r}
#heart disease
ggplot(final.data[which(final.data$Cause.Name=="Heart disease"),], aes(Year, rate)) +
  ggtitle("Year vs Mortality Rate due to Heart Disease") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Year") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="green", fill="lightgreen") 
#Cancer
ggplot(final.data[which(final.data$Cause.Name=="Cancer"),], aes(Year, rate)) +
  ggtitle("Year vs Mortality Rate due to Cancer") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Year") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="green", fill="lightgreen") 
#Stroke
ggplot(final.data[which(final.data$Cause.Name=="Stroke"),], aes(Year, rate)) +
  ggtitle("Year vs Mortality Rate due to Stroke") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Year") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="green", fill="lightgreen") 
#CLRD
ggplot(final.data[which(final.data$Cause.Name=="CLRD"),], aes(Year, rate)) +
  ggtitle("Year vs Mortality Rate due to CLRD") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Year") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="red", fill="red")
#Diabetes
ggplot(final.data[which(final.data$Cause.Name=="Diabetes"),], aes(Year, rate)) +
  ggtitle("Year vs Mortality Rate due to Diabetes") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Year") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="red", fill="red")
#Alzheimer's
ggplot(final.data[which(final.data$Cause.Name=="Alzheimer's disease"),], aes(Year, rate)) +
  ggtitle("Year vs Mortality Rate due to Alzheimer's Disease") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Year") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="red", fill="red")
#Kidney Disease
ggplot(final.data[which(final.data$Cause.Name=="Kidney disease"),], aes(Year, rate)) +
  ggtitle("Year vs Mortality Rate due to Kidney Disease") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Year") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="red", fill="red")
#Flu/pneumonia
ggplot(final.data[which(final.data$Cause.Name=="Influenza and pneumonia"),], aes(Year, rate)) +
  ggtitle("Year vs Mortality Rate due to Influenza and Pneumonia") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Year") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="green", fill="lightgreen") 

#Health Expenditures
#heart disease
ggplot(final.data[which(final.data$Cause.Name=="Heart disease"),], aes(health, rate)) +
  ggtitle("Health Expenditure vs Mortality Rate due to Heart Disease") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Health Expenditure in Billions") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="green", fill="lightgreen") 
#Cancer
ggplot(final.data[which(final.data$Cause.Name=="Cancer"),], aes(health, rate)) +
  ggtitle("Health Expenditure vs Mortality Rate due to Cancer") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Health Expenditure in Billions") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="green", fill="lightgreen") 
#Stroke
ggplot(final.data[which(final.data$Cause.Name=="Stroke"),], aes(health, rate)) +
  ggtitle("Health Expenditure vs Mortality Rate due to Stroke") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Health Expenditure in Billions") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="green", fill="lightgreen") 
#CLRD
ggplot(final.data[which(final.data$Cause.Name=="CLRD"),], aes(health, rate)) +
  ggtitle("Health Expenditure vs Mortality Rate due to CLRD") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Health Expenditure in Billions") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="red", fill="red")
#Diabetes
ggplot(final.data[which(final.data$Cause.Name=="Diabetes"),], aes(health, rate)) +
  ggtitle("Health Expenditure vs Mortality Rate due to Diabetes") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Health Expenditure in Billions") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="red", fill="red")
#Alzheimer's
ggplot(final.data[which(final.data$Cause.Name=="Alzheimer's disease"),], aes(health, rate)) +
  ggtitle("Health Expenditure vs Mortality Rate due to Alzheimer's Disease") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Health Expenditure in Billions") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="red", fill="red")
#Kidney Disease
ggplot(final.data[which(final.data$Cause.Name=="Kidney disease"),], aes(health, rate)) +
  ggtitle("Health Expenditure vs Mortality Rate due to Kidney Disease") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Health Expenditure in Billions") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="red", fill="red")
#Flu/pneumonia
ggplot(final.data[which(final.data$Cause.Name=="Influenza and pneumonia"),], aes(health, rate)) +
  ggtitle("Health Expenditure vs Mortality Rate due to Influenza and Pneumonia") +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line() +
  geom_point() +
  scale_color_discrete("Cause of Death") +
  xlab("Health Expenditure in Billions") +
  ylab("Mortality Rate per 100k") +
  geom_smooth(method = lm, level = 0.95, color="green", fill="lightgreen") 
```
